Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-06-05 paper/paper.tex
done with paper text

\vspace{0.5em}\tiny{$\dagger$ These authors have equally contributed to this article.}
}
\def\corrAuthor{Srinivas Gorur-Shandilya \vspace{1mm}}
%\def\corrAddress{Marder Lab, Brandeis University, Biology Department and Volen Center for Complex Systems, Waltham, MA, USA}
\def\corrAddress{Volen Center and Biology Department}
\def\corrEmail{\texttt{srinivasgs@brandeis.edu}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		INTRODUCTION & OVERVIEW
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
%									 DESIGN GOALS
% 
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

"xolotl" is designed to be easy-to-use and richly featured and that is fast enough to use in everyday research. Our focus was on designing a simulator of conductance-based neurons and networks of these neurons; and simulating arbitrary dynamical systems [cite DynaSim and others] is therefore beyond the scope of this software. The software is designed to be used from within "MATLAB", a scientific programming language popular amongst neuroscientists and engineers for pedagogy and research. Our goal is to make "xolotl" completely usable entirely within MATLAB, and all parameters of models can be be inspected and edited before running the model. Several features of the simulator are designed to be easily extensible, and the simulator appears in the "MATLAB" workspace as a first-class native object, and is fully compatible with the large library of toolboxes that a scientific programming environment like "MATLAB" provides.

@@

"MATLAB" provides a high level programmatic and graphical interface for implementing, manipulating, and visualizing models without sacrificing the enhancements of the underlying "C++" code.

\paragraph{Library of network components.} "xolotl" comes packaged with hundreds of pre-existing components (viz. compartments, synapses, conductances, and controllers) which greatly simplify the task of constructing model neurons. Compartments represent cylindrical sections of membrane with a membrane potential. Synapses connect two compartments together by introducing a current in the post-synaptic neuron that depends on the state of the presynaptic neuron. Conductances represent populations of ion channels in a compartment that produce transmembrane currents. Controllers integrate a signal from a conductance or synapse and change the properties of the conductance or synapse in response.

\paragraph{Object-oriented.}
Models are specified by adding compartments and synapses to the "xolotl" object. Conductances are added to compartments and controllers can be added to conductances or synapses. This modular structure recapitulates the biophysics of the Hodgkin-Huxley formalism and obviates the need to explicitly write out equations, which in "xolotl" are contained within "C++" header files.

@@

\paragraph{Fast.}
"xolotl" is fast because all time-intensive code is written in "C++". While automated "C++" transpiling from "MATLAB" using the proprietary "codegen" can drastically improve performance over loops through strong typing and memory pre-allocation, supervenience of "MATLAB" over "C++" prevents efficient use of features, such as passing by reference and object-oriented programming. "C++" provides speed improvements beyond the benefits of translating "MATLAB" features into statically-typed code. For this reason, "cpplab" has been designed to provide an interface for constructing, transpiling, and compiling "C++" code to be called from within "MATLAB".

@@

The focus on ease-of-use and fast simulations means some features were intentionally neglected in the streamlining process.

\paragraph{Limited to conductance-based models.} "xolotl" has been developed specifically for conductance-based models. It does not currently support rate- or current-based models.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		USAGE EXAMPLES
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this section, we describe several uses of "xolotl" for common situations.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{\texttt{xolotl} can quickly set up and simulate conductance-based models. (A) Cartoon of a Hodgkin-Huxley single-compartment neuron model with fast sodium, delayed rectifier, and leak currents. (B) Code snippet in \texttt{MATLAB} used to implement D, F-I. (C) \texttt{xolotl} schematic displayed in the \texttt{MATLAB} command prompt. (D) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium, blue is delayed rectifier, red is leak). (E) Firing rate-input relation displaying firing rate as a function of injected current current. (F-G) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (H-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating var

>>>>

\vspace{0.5em}\tiny{$\dagger$ These authors have equally contributed to this article.}
}
\def\corrAuthor{Srinivas Gorur-Shandilya \vspace{1mm}}
%\def\corrAddress{Marder Lab, Brandeis University, Biology Department and Volen Center for Complex Systems, Waltham, MA, USA}
\def\corrAddress{..}
\def\corrEmail{\texttt{srinivasgs@brandeis.edu}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		INTRODUCTION & OVERVIEW
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		DESIGN GOALS
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

"xolotl" is designed to be easy-to-use and richly featured and that is fast enough to use in everyday research. Our focus was on designing a simulator of conductance-based neurons and networks of these neurons; and simulating arbitrary dynamical systems is therefore beyond the scope of this software. The software is designed to be used from within "MATLAB", a scientific programming language popular amongst neuroscientists and engineers for pedagogy and research. Our goal is to make "xolotl" completely usable entirely within MATLAB, and all parameters of models can be be inspected and edited before running the model. Several features of the simulator are designed to be easily extensible, and the simulator appears in the "MATLAB" workspace as a first-class native object, and is fully compatible with the large library of toolboxes that a scientific programming environment like "MATLAB" provides.

@@

\paragraph{A rich library of network components.} "xolotl" comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and controllers) which are used to construct model neurons and networks. Compartments represent cylindrical sections of membrane with a single membrane potential, and can represent either entire cells or parts of cells. Synapses connect two compartments together by introducing a current in the post-synaptic neuron that depends on the voltage of the presynaptic neuron. Conductances represent populations of ion channels in a compartment that produce transmembrane currents. Controllers represent homeostatic mechanisms that integrate a Calcium-dependent signal from the compartment they are in and  change the properties of a conductance or synapse in response.

\paragraph{Object-oriented.}
"xolotl" is designed to mirror the nested and hierarchical structure of networks and neurons.  Biological neuronal networks consist of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that affect its electrical behavior. Similarly, a "xolotl" model can contain a set of compartments that can represent individual neurons. Each compartment can contain a set of conductances. All objects in the "xolotl" model tree (compartments, conductances, etc.) are bonafide "MATLAB" objects that have their own type, properties and associated methods. This object-oriented programming paradigm naturally represents the hierarchical structure of biological neurons and networks.

@@

@@

The focus on ease-of-use and fast simulations means some features were intentionally neglected in the design process.

\paragraph{Limited to conductance-based models.} "xolotl" has been developed specifically for conductance-based models. It does not currently support rate- or current-based models, or arbitrary dynamical systems.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		USAGE EXAMPLES
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this section, we illustrate how "xolotl" can be used to generate a variety of models and simulate them. These examples have been chosen to illustrate various features of "xolotl", and are intended to serve as a template that researchers and educators can use for their own needs.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{\texttt{xolotl} can quickly set up and simulate conductance-based models. (A) Cartoon of a Hodgkin-Huxley single-compartment neuron model with fast sodium, delayed rectifier, and leak currents. (B) Code snippet in \texttt{MATLAB} used to implement D, F-I. (C) \texttt{xolotl} schematic displayed in the \texttt{MATLAB} command prompt. (D) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium, blue is delayed rectifier, red is leak). (E) Firing rate-input relation displaying firing rate as a function of injected current current. (F-G) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (H-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables.}
	\label{fig:figurehh}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) Cartoon of a cell with delayed rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998} with experimentally-fixed voltage. (B) Structure of \texttt{xolotl} object in A. (C) Code snippet depicting integration under voltage clamp. (D-E) Current response to steps in voltage from a holding potential of $V_m = -60$ mV. (F) Current-voltage relation of the steady-state current ($t = 400$ ms) indicating a reversal potential of $E = -80$ mV and no inactivation. (G) Conductance-voltage relation at steady-state takes the form of a sigmoid. (H) Sigmoids $m$ fit to the model as $m^n$ data indicating that $n=4$ is the best fit. (I) Goodness of fit vs. exponent $n$, suggesting $n=4$ as the best fit to the data.}
	\label{fig:figureclamp}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_network}
	\caption{Simulating a network of conductance-based model neurons. (A) Diagram of a network model of the pyloric rhythm in the crustacean stomatogastric ganglion (Prinz \textit{et al.} 2004). (B) Each neuron is modeled as a single compartment with 7-8 intrinsic conductances and 1-3 post-synaptic conductances. (C) \texttt{xolotl} implements conductances as fields of compartments and synapses as connections between compartments. (D-F) Simulated voltage trace of a model network for the three compartments. (G) Time series of synaptic currents in the simulated network can be obtained from the integration.}
	\label{fig:figurenetwork}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_integral_control}
	\caption{Simulating neurons under homeostatic regulation. (A) Cartoon of a model neuron (\cite{liuModelNeuronActivitydependent1998} with integral control (\cite{olearyCorrelationsIonChannel2013}. (B) Hierarchical structure of a neuronal network considers controllers as components of compartments which act on conductances. (C) \texttt{xolotl} implements controllers as properties of conductances and synapses. (D) Calcium sensors change maximal conductances to move a neuron from quiescence to a bursting state. (E) Voltage trace shows regular bursting activity after integral control.}
	\label{fig:figureintegralcontrol}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/puppeteer_screenshot}
	\caption{Using the GUI to manipulate neuron parameters. (A) Real-time output of the \texttt{plot} function displaying voltage (colored) and intracellular calcium (black) traces of a bursting neuron model (\cite{prinzAlternativeHandtuningConductancebased2003, prinzSimilarNetworkActivity2004}. Colors indicate the dominant current. (B) Sliders control the maximal conductances, which updates on the figure.}
	\label{fig:puppeteerscreenshot}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{\texttt{xolotl} benchmarked against \texttt{DynaSim} and \texttt{NEURON}. (A) Ratio of 'simulated' time to runtime (speed factor) and accuracy, measured by spike train coincidence plotted against decreasing time-resolution. (B) Speed factor for models at increasing simulation times. (C) Speed factor over number of compartments.}
	\label{fig:figurebenchmark}
\end{figure}

\end{document}
